Making Deep Q-learning Methods Robust to Time Discretization

### 2019-01-23 paper/icml_drau/advup.tex
.

The discretized $Q$-function rewrites as
\begin{equation}
	Q^\pi_\deltat(s, a) = V^\pi_\deltat(s) + \deltat A^\pi_\deltat(s, a).
	\label{eq:reparam_q_pi}
\end{equation}
A natural way to approximate $V^\pi_\deltat$ and $A^\pi_\deltat$ is to apply
\emph{Sarsa} or $Q$-learning to a reparameterized $Q$-function approximator
\begin{equation}
	Q_\Theta(s, a) \deq V_{\theta}(s) + \deltat A_{\psi}(s, a).
\end{equation}
with $\Theta \deq (\theta, \psi)$. (To avoid cumbersome notation, 
parameter indices are dropped in what follows.)  At initialization, if both $V_{\theta}$ and
$A_{\psi}$ are initialized independently of $\deltat$, this parameterization
provides reasonable scaling of the contribution of actions versus states
in $Q$.
Our goal is for $V_\theta$ to approximate $V^\pi_\deltat$ and for
$A_{\psi}$ to approximate $A^\pi_\deltat$.

%\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
Still, this reparameterization does not, on its own, guarantee that $A$
correctly approximates $A^\pi_\deltat$ if
$Q$ approximates $Q^\pi_\deltat$% \footnote{
% 	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
% 	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
% 	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is
% 	the same. \TODO{still unclear why we'd care about identifiability}
%}
.
Indeed, for any given pair $(V,\,A)$, the pair $(V(s) -
f(s),\,A(s,a)+f(s)/\deltat)$ (for an arbitrary $f$)
% \begin{align}
% 	\tilde{V}(s) &= V(s) - f(s),\qquad
% 	\tilde{A}(s, a) = A(s, a) + \frac{f(s)}{\deltat}
% \end{align}
yields the exact same $Q$ function. This new $A$ still defines %the same $Q$ and
the same ranking of actions, yet this phenomenon might cause
numerical problems or instability of $A$ when $\deltat\to 0$, and prevents direct
interpretation of the learned $A$.
% Consequently, $Q$ can correctly approximate
% $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
To enforce identifiability of $A$, one must enforce the consistency equation
\begin{equation}
	Q^\pi_\deltat(s, \pi(s)) - V^\pi_\deltat(s) = 0
\end{equation}
on the approximate $A$ and $V$. This translates to
\begin{equation}
	A(s, \pi(s)) = 0.
\end{equation}
With this additional constraint, if $Q = Q^\pi_\deltat$, then $A =
A^\pi_\deltat$ and $V = V^\pi_\deltat$: indeed \TODO{keep or  leave as
an exercise?}
\begin{align}
	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}
		    %\\ &
		    = A(s, a).
\end{align}
In the spirit of~\cite{dueling_nets}, this condition is enforced by
parameterizing $A$ through another function $\bar{A}$ and setting
\begin{equation}
\label{eq:Aparam}
	A(s, a) \deq \bar{A}(s, a) - \bar{A}(s, \pi(s)).
\end{equation}

@@

\begin{theorem}
Under suitable smoothness assumptions, $A^\pi_\deltat(s,a)$ has a limit
$A^\pi(s,a)$ when $\deltat\to 0$. The limit is informative about
actions: namely, if a policy $\pi'$ strictly dominates $\pi$, 
%satisfies $V^{\pi'}>V^{\pi}$ for all states,
then
$A^\pi(s,\pi'(s))>0$ for some state $s$.
\end{theorem}

@@

Thus, for discrete actions we can obtain a temporally consistent
exploration scheme by taking
$z_\deltat$ to be a discretization of an $|\mathcal{A}|$-dimensional
continuous OU process, and setting
\begin{equation}
  \pi^\text{explore}(s_t, z_t)\deq \text{argmax}_{a'} \left(A(s, a') + z_t[a']\right)
\end{equation}
where $z_t[a']$ denotes the $a'$-th component of $z_t$. Namely, we
perturb advantage values by a random process before selecting an action. The resulting scheme
converges in continuous time to a nontrivial exploration scheme.

@@

Maximization over actions in $\pi$ is implemented exactly for discrete actions,
and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A(s,
\pi_\phi(s))$ for continuous actions. This point is similar to
\cite{ddpg}.

Eq.~\eqref{eq:max_A}
is directly verified by $A_{\psi}$, owing to the reparametrization
$A_\psi(s, a) = \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s))$, described
in~\ref{subsec:reparam}.  To approximately verify~\eqref{eq:bellman_A}, the
corresponding squared residual is minimized by an approximate gradient descent.
The update equations when learning from a transition $(s, a, r, s')$, either from
an exploratory trajectory or from a replay buffer~\cite{dqn}, are
\begin{align}
	\delta Q_\deltat &\leftarrow A(s, a)\, \deltat - \left(r
	\,\deltat + \gamma^{\deltat}\, V(s') - V(s)\right)
	\label{eq:approx_deltaQ}\\
	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat
	\,\partial_{\theta} V(s) \,\frac{\delta Q_\deltat}{\deltat}
	\label{eq:approx_bellman_V}\\
	\psi_\deltat &\leftarrow \psi_\deltat + \eta^A_\deltat
	\,\partial_{\psi} A(s, a) \,\frac{\delta Q_\deltat}{\deltat}.
	\label{eq:approx_bellman_A}
\end{align} 
where the $\eta$'s are learning rates.
Appropriate scalings for the learning rates $\eta^V_\deltat$ and
$\eta^A_\deltat$ in term of $\deltat$ to obtain a well defined continuous
limit are derived next.
% As we will see in the following section,
% to obtain a well defined continuous time limit, the learning rates can be rewritten as
% $\eta^i_\deltat = \alpha^i \deltat$, where $\alpha^1$ and $\alpha^2$ are independant
% of $\deltat$.

>>>>

The discretized $Q$-function rewrites as
\begin{equation}
	Q^\pi_\deltat(s, a) = V^\pi_\deltat(s) + \deltat A^\pi_\deltat(s, a).
	\label{eq:reparam_q_pi}
\end{equation}
A natural way to approximate $V^\pi_\deltat$ and $A^\pi_\deltat$ is to apply
\emph{Sarsa} or $Q$-learning to a reparameterized $Q$-function approximator
\begin{equation}
	Q_\Theta(s, a) \deq V_{\theta}(s) + \deltat A_{\psi}(s, a).
\end{equation}
with $\Theta \deq (\theta, \psi)$. At initialization, if both $V_{\theta}$ and
$A_{\psi}$ are initialized independently of $\deltat$, this parameterization
provides reasonable scaling of the contribution of actions versus states
in $Q$.
Our goal is for $V_\theta$ to approximate $V^\pi_\deltat$ and for
$A_{\psi}$ to approximate $A^\pi_\deltat$.

%\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
Still, this reparameterization does not, on its own, guarantee that $A$
correctly approximates $A^\pi_\deltat$ if
$Q_\Theta$ approximates $Q^\pi_\deltat$% \footnote{
% 	Ensuring that $A$ does not differ from $A^\pi_\deltat$ by a state dependent function only affects
% 	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
% 	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is
% 	the same. \TODO{still unclear why we'd care about identifiability}
%}
.
Indeed, for any given pair $(V_\theta,\,A_\psi)$, the pair $(V_\theta(s) -
f(s),\,A_\psi(s,a)+f(s)/\deltat)$ (for an arbitrary $f$)
% \begin{align}
% 	\tilde{V}(s) &= V(s) - f(s),\qquad
% 	\tilde{A}(s, a) = A(s, a) + \frac{f(s)}{\deltat}
% \end{align}
yields the exact same $Q_\Theta$ function. This new $A_\psi$ still defines %the same $Q$ and
the same ranking of actions, yet this phenomenon might cause
numerical problems or instability of $A_\psi$ when $\deltat\to 0$, and prevents direct
interpretation of the learned $A_\psi$.
% Consequently, $Q$ can correctly approximate
% $Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
To enforce identifiability of $A_\psi$, one must enforce the consistency equation
\begin{equation}
	Q^\pi_\deltat(s, \pi(s)) - V^\pi_\deltat(s) = 0
\end{equation}
on the approximate $A_\psi$ and $V_\theta$. This translates to
\begin{equation}
	A_\psi(s, \pi(s)) = 0.
\end{equation}
With this additional constraint, if $Q_\Theta = Q^\pi_\deltat$, then $A_\psi =
A^\pi_\deltat$ and $V_\theta = V^\pi_\deltat$: indeed \TODO{keep or  leave as
an exercise?}
\begin{align}
	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
		    &= \frac{Q_\Theta(s,a) - Q_\Theta(s, \pi(s))}{\deltat}
		    %\\ &
		    = A_\psi(s, a).
\end{align}
In the spirit of~\cite{dueling_nets}, instead of directly parameterizing $A_\psi$,
we define a parametric function $\bar{A}_\psi$ (typically a neural network),
and use $\bar{A}_\psi$ to define $A_\psi$ as
\begin{equation}
\label{eq:Aparam}
	A_\psi(s, a) \deq \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s)).
\end{equation}
% Rq Harsh
With this definition, $A_\psi$ directly verifies the normalization condition.

@@

\begin{theorem}
Under suitable smoothness assumptions, $A^\pi_\deltat(s,a)$ has a limit
$A^\pi(s,a)$ when $\deltat\to 0$. The limit keeps information about
actions: namely, if a policy $\pi'$ strictly dominates $\pi$, 
%satisfies $V^{\pi'}>V^{\pi}$ for all states,
then
$A^\pi(s,\pi'(s))>0$ for some state $s$.
\end{theorem}

@@

Thus, for discrete actions we can obtain a temporally consistent
exploration scheme by taking
$z_\deltat$ to be a discretization of an $|\mathcal{A}|$-dimensional
continuous OU process, and setting
\begin{equation}
  \pi^\text{explore}(s^k_\deltat, z^k_\deltat)\deq \text{argmax}_{a'} \left(A_\psi(s^k_\deltat, a') + z^k_\deltat[a']\right)
\end{equation}
where $z^k_\deltat[a']$ denotes the $a'$-th component of $z^k_\deltat$. Namely,
we perturb advantage values by a random process before selecting an action. The
resulting scheme converges in continuous time to a nontrivial exploration
scheme.

@@

Maximization over actions in $\pi$ is implemented exactly for discrete actions,
and approximated by a policy neural network $\pi_\phi(s)$, trained to maximize $A_\psi(s,
\pi_\phi(s))$ for continuous actions. This point is similar to
\cite{ddpg}.

Eq.~\eqref{eq:max_A}
is directly verified by $A_{\psi}$, owing to the reparametrization
$A_\psi(s, a) = \bar{A}_\psi(s, a) - \bar{A}_\psi(s, \pi(s))$, described
in~\ref{subsec:reparam}.  To approximately verify~\eqref{eq:bellman_A}, the
corresponding squared residual is minimized by an approximate gradient descent.
The update equations when learning from a transition $(s, a, r, s')$, either from
an exploratory trajectory or from a replay buffer~\cite{dqn}, are
\begin{align}
	\delta Q_\deltat &\leftarrow A_\psi(s, a)\, \deltat - \left(r
	\,\deltat + \gamma^{\deltat}\, V_\theta(s') - V_\theta(s)\right)
	\label{eq:approx_deltaQ}\\
	\theta_\deltat &\leftarrow \theta_\deltat + \eta^V_\deltat
	\,\partial_{\theta} V_\theta(s) \,\frac{\delta Q_\deltat}{\deltat}
	\label{eq:approx_bellman_V}\\
	\psi_\deltat &\leftarrow \psi_\deltat + \eta^A_\deltat
	\,\partial_{\psi} A_\psi(s, a) \,\frac{\delta Q_\deltat}{\deltat}.
	\label{eq:approx_bellman_A}
\end{align} 
where the $\eta$'s are learning rates.
Appropriate scalings for the learning rates $\eta^V_\deltat$ and
$\eta^A_\deltat$ in term of $\deltat$ to obtain a well defined continuous
limit are derived next.
% As we will see in the following section,
% to obtain a well defined continuous time limit, the learning rates can be rewritten as
% $\eta^i_\deltat = \alpha^i \deltat$, where $\alpha^1$ and $\alpha^2$ are independant
% of $\deltat$.

### 2019-01-23 paper/icml_drau/intro.tex
.

Standard approaches based on estimation of state-action
value functions, such as
\emph{Deep $Q$-learning} (DQN,~\citealt{dqn}) and \emph{Deep deterministic policy
gradient} (DDPG,~\citealt{ddpg}) are not at all robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
Intuitively, as the discretization timestep
decreases, the effect of individual actions on the total return decreases
too: $Q^*(s,a)$ is the value of playing action $a$ then playing optimally,
and if $a$ is only maintained for a very short time its advantage over
other actions will be accordingly small. (This occurs even with a
suitably adjusted decay factor $\gamma$.)

\TODO{
If the discretization timestep becomes infinitesimal, the effect of every
individual action vanishes: there is no continuous-time $Q$-function
(Thm.~\ref{th:q-cont}). These statements can be fully formalized in the framework
of continuous-time reinforcement learning (Sec.~\ref{sec:framework})
\cite{cont_rl, adv_upd}.

We focus on continuous time because this leads to a clear theoretical framework,
but our observations make sense in any setting in which the value
results from taking a large number of small individual actions. Our
results suggest standard $Q$-learning will fail in such settings without a
delicate balance of hyperparameter scalings and reparameterizations.}
%the
%advantage of any single individual action is small compared to the total
%return.%, for instance, long time horizon problems.

>>>>

Standard approaches based on estimation of state-action
value functions, such as
\emph{Deep $Q$-learning} (DQN,~\citealt{dqn}) and \emph{Deep deterministic policy
gradient} (DDPG,~\citealt{ddpg}) are not at all robust to changes in time discretization. This is shown experimentally in Sec.~\ref{sec:exp}. 
Intuitively, as the discretization timestep
decreases, the effect of individual actions on the total return decreases
too: $Q^*(s,a)$ is the value of playing action $a$ then playing optimally,
and if $a$ is only maintained for a very short time its advantage over
other actions will be accordingly small. (This occurs even with a
suitably adjusted decay factor $\gamma$.)
If the discretization timestep becomes infinitesimal, the effect of every
individual action vanishes: there is no continuous-time $Q$-function
(Thm.~\ref{th:q-cont}), hence the poor performance of $Q$-learning with
small time steps. These statements can be fully formalized in the framework
of continuous-time reinforcement learning (Sec.~\ref{sec:framework})
\cite{cont_rl, adv_upd}.

We focus on continuous time because this leads to a clear theoretical framework,
but our observations make sense in any setting in which the value
results from taking a large number of small individual actions. Our
results suggest standard $Q$-learning will fail in such settings without a
delicate balance of hyperparameter scalings and reparameterizations.
%the
%advantage of any single individual action is small compared to the total
%return.%, for instance, long time horizon problems.
